Emergent dual scaling of riverine biodiversity

### 2021-03-23 coverletter.Rmd
Update coverletter.Rmd

---
output: pdf_document
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
---

@@

In the Anthropocene, it is more imperative than ever to identify ecological factors that underpin diverse ecological communities in nature. Much attention has been paid to the ubiquitous power-law scaling of species richness and area, in which species richness increases with area in a decelerating way. Ecologists have used this ecological *law* to inform many conservation strategies, including marine and terrestrial protected areas that currently encompass > 30 million km^2^ globally. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because 'scale-invariant' complexity of ecosystem structure may drive ecological dynamics in space. Branching river networks are an outstanding example of such complex ecosystems, where geometric ecosystem structure (e.g., branching patterns) remains constant across a wide range of spatial scales.

Here, we used theory and analysis of large fish community data to show that **ecosystem complexity of branching networks (measured as the prevalence of branching) plays a crucial role in scaling metacommunity-level biodiversity in rivers (**$\gamma$ **diversity); ecosystem size and other macro-scale drivers (e.g., climates and land use patterns) alone were insufficient to describe biodiversity patterns in complex river networks.** Our theory predicted that, regardless of simulated species' traits, more branched 'complex' river networks supported greater species richness due to increased habitat heterogeneity. This relationship had a characteristic of power-law, revealing the overlooked scaling relationship in complex branching networks. Importantly, the complexity effect was comparable to the size effect - hence, riverine biodiversity should follow a dual scaling law along the two orthogonal dimensions of ecosystem structure, i.e., size and scale-invariant complexity.

We confirmed the theoretical predictions using extensive fish community data, which included 180 replicates of metacommunities (i.e., watersheds) across two geographic regions (Japan and the US). The dual scaling of riverine fish diversity has consistently emerged across the study regions despite hosting different fish fauna with distinct evolutionary histories. This result agrees elegantly with theoretical predictions, in which species' traits played a minor role in shaping the dual power-law scaling. In particular, the effect of ecosystem complexity was striking in its magnitude. A unit increase of ecosystem complexity (the prevalence of branching) resulted in an increase of species diversity that would require 1911 km^2^ in the areal area.

>>>>

---
output: word_document
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
---

![](images/uncgreensboro_DoB_h_pms-c.png){width="200"}

@@

In the Anthropocene, it is more imperative than ever to identify ecological factors that underpin diverse ecological communities in nature. Much attention has been paid to the ubiquitous power-law scaling of species richness and area, in which species richness increases with area in a decelerating way. Ecologists have used this ecological *law* to inform many conservation strategies, including marine and terrestrial protected areas that currently encompass \> 30 million km^2^ globally. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because 'scale-invariant' complexity of fractal ecosystem structure may drive ecological dynamics in space. Branching river networks are an outstanding example of such complex ecosystems, where geometric ecosystem structure (e.g., branching patterns) remains constant across a wide range of spatial scales.

Here, we used theory and analysis of large fish community data to show that **ecosystem complexity of branching networks (measured as the prevalence of branching) plays a crucial role in scaling metacommunity-level biodiversity in rivers (**$\gamma$ **diversity); ecosystem size and other macro-scale drivers (e.g., climates and land use patterns) were insufficient to describe riverine biodiversity patterns.** Our theory predicted that, regardless of simulated species' traits, more branched 'complex' river networks supported greater species richness due to increased habitat heterogeneity. This relationship had a characteristic of power-law, revealing the overlooked scaling relationship in complex branching networks. Importantly, the complexity effect was comparable to the size effect - hence, riverine biodiversity should follow a dual scaling law along the two orthogonal dimensions of ecosystem structure, i.e., size and scale-invariant complexity.

We confirmed the theoretical predictions using extensive fish community data, which included 180 replicates of metacommunities (i.e., watersheds) across two geographic regions (Japan and the US). The dual scaling of riverine fish diversity has consistently emerged across the study regions despite hosting different fish fauna with distinct evolutionary histories. This result agrees elegantly with our theoretical predictions, in which species' traits played a minor role in shaping the dual power-law scaling. Further, the effect of ecosystem complexity was striking in its magnitude. A unit increase of ecosystem complexity (the prevalence of branching) resulted in an increase of species diversity that would require 1911 km^2^ in the areal area.
